---
title: "Trajectories against moving boundaries"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Trajectories against moving boundaries}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.2)
```

```{r setup}
library(ggplot2)
library(ggboundary)
```

## The idea

A boundary is not always a constant. Doughnut Economics and the planetary-boundaries wheel draw the safe threshold as a fixed line, but a system's safe threshold often moves with its own capacity: a tourism economy's carrying capacity rises as it builds room stock and can be lowered again by policy, a reserve buffer's adequacy shifts with the risk it faces. `ggboundary` draws the boundary as a series so that movement is visible, and shades where the trajectory overshoots it.

## `geom_boundary()`

```{r}
df <- data.frame(
  year = 2016:2025,
  hhi  = c(38, 40, 41, 43, 45, 47, 49, 52, 54, 55),
  cap  = c(50, 50, 51, 52, 52, 50, 48, 47, 47, 48)
)

ggplot(df) +
  geom_boundary(df, "year", "hhi", "cap") +
  labs(y = "concentration (HHI)", x = NULL) +
  theme_minimal()
```

The dashed line is the boundary, moving. Red marks overshoot, green marks slack.

## The doughnut, as the single-instant snapshot

```{r, fig.width = 5, fig.height = 5}
d <- data.frame(dim = c("water", "food", "energy", "income", "health"),
                v = c(0.2, 0.5, 0.8, 0.45, 0.7))
doughnut(d, "dim", "v")
```

## The conch, as the expressive companion

```{r, fig.width = 5, fig.height = 5}
conch(df$year, df$hhi, df$cap, label_fmt = round,
      title = "The same series, wound through time")
```

Use the conch to make the boundary idea memorable, not to read values off. When precision matters, the Cartesian `geom_boundary()` above is the tool.
